Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2025-02-02 ms.tex
edited app D

In the following, we assume aneuploid cells are tolerant ($\Delta_a T^* \ll -1$) and frequent enough to affect the evolution of drug resistance (i.e. $u\lambda_a \gg \max{(-\Delta_a, 1/T^*)}$). 
We define the proliferation time $\tau_a^p$  to be the expected time for the number of resistant cells to grow to the initial tumor size $N$.
The number of rescue lineages generated by the sensitive population is given by  \cref{eq:twosteplineage} (see \Cref{ExpectedNumberRescueLineages}),
\begin{equation*}
n_1\left(\infty\right)=\frac{uv\lambda_s\lambda_aNp_m}{\abs{\Delta_s}\abs{\Delta_a}}+\frac{v\lambda_sNp_m}{\abs{\Delta_s}}=\frac{N}{N_a^*}+\frac{N}{N_m^*} .
\end{equation*}
We use \cref{eq:N_a,eq:N_m} to write $N_a^*=\frac{\abs{\Delta_s}\; \abs{\Delta_a}}{uv\lambda_s\lambda_ap_m}$ and $N_m^*=\frac{\abs{\Delta_s}}{v\lambda_sp_m}$.

We distinguish between two scenarios for the proliferation time. 
In the first scenario, $N\ll N_a^*\ll N_m^*$, and we have at most one lineage that rescues the cancer cell population.
As a result, the recurrence time is given by~\citep{avanzini2019cancer}
\begin{align}\label{meanproliferationtime}
\tau_a^r&\approx\tau_a+\frac{\log \left(p_mN\right)}{\Delta_m}.
\end{align}
The factor $p_m$ in the second term of \cref{meanproliferationtime} occurs because the lineage is conditioned to survive stochastic extinction and the time to reach $N$ is shorter compared to the scenario where the lineage is not conditioned to survive stochastic extinction~\citep{orr2014population,smith1974hitch}. % change 35

In the second scenario, $N\gg N_m^*$, and the sensitive population produces a large number of rescue lineages in a short period of time.
As a result, the recurrence time is obtained by solving the following system of ordinary differential equations (ODEs),
\begin{equation}\label{detODE}
\begin{aligned}
\frac{ds}{dt}&=\Delta_ss,\\
\frac{da}{dt}&=\Delta_aa+u\lambda_ss,\\
\frac{dm}{dt}&=\Delta_mm+v\lambda_aa+v\lambda_ss.
\end{aligned}
\end{equation}
Solving the system of ODEs for initial condition $\left(s(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain:
\begin{equation*} \label{m_t}
m(t) = \frac{N u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m t} - \e^{\Delta_a t}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s} \right] + N v \lambda_s \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s}.
\end{equation*}
We obtain $\tau_a^r$ by solving the above equation for time $t=\tau_a^r$ at which the number of mutant cells reaches $N$, that is, we solve $m\left(\tau_a^r\right)=N$,
\begin{equation}\label{eq:t2det2024}
1 = \frac{u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_a \tau_a^r}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s} \right] + v \lambda_s \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s}.
\end{equation}

\eqref{eq:t2det2024} is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that the final tumor size $N$ is very large, and that since the sensitive and aneuploid cells have negative growth rates, almost all of the growth needed to rebound to size $N$ will be by mutant cells. 
Mathematically, this corresponds to $\e^{\Delta_m \tau_a^r}$ being much larger than the other exponential terms,  $\e^{\Delta_s \tau_a^r}$ and $\e^{\Delta_a \tau_a^r}$. 
\eqref{eq:t2det2024} can then be rewritten as:
\[
	1 \approx \frac{v \lambda_s e^{\Delta_m \tau_a^r}}{\Delta_m - \Delta_s} 
	\left(\frac{u \lambda_a}{\Delta_m - \Delta_a} + 1\right).
\]
Assuming that rate of aneuploidy is not so high that it overwhelms the aneuploids' growth disadvantage relative to the mutants (i.e., that $u \lambda_a \ll \Delta_m - \Delta_a$), we can neglect the first term in parentheses above, yielding:
\begin{equation}\label{eq:t2det}
\tau_a^r\approx\frac{1}{\Delta_m}\log{\left(\frac{\Delta_m-\Delta_s}{v\lambda_s}\right)}.
\end{equation}
We emphasize that in this case the population is large enough to produce mutants without needing to pass through aneuploidy ($N \gg N_m^*$), and so  the terms arising from the evolutionary trajectory \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{mutant} do not contribute to the above approximation. %change 9

>>>>

In the following, we assume aneuploid cells are tolerant ($\Delta_a T^* \ll -1$) and frequent enough to affect the evolution of drug resistance ($u\lambda_a \gg \max{(-\Delta_a, 1/T^*)}$). 
We define the proliferation time $\tau_a^p$  to be the expected time for the number of resistant cells to grow to the initial tumor size $N$.
The number of rescue lineages generated by the sensitive population is given by  \cref{eq:twosteplineage} (see \Cref{ExpectedNumberRescueLineages}),
\begin{equation*}
n_1\left(\infty\right)=\frac{uv\lambda_s\lambda_aNp_m}{\abs{\Delta_s}\abs{\Delta_a}}+\frac{v\lambda_sNp_m}{\abs{\Delta_s}}=\frac{N}{N_a^*}+\frac{N}{N_m^*} .
\end{equation*}
We use \cref{eq:N_a,eq:N_m} to write $N_a^*=\frac{\abs{\Delta_s}\; \abs{\Delta_a}}{uv\lambda_s\lambda_ap_m}$ and $N_m^*=\frac{\abs{\Delta_s}}{v\lambda_sp_m}$.

We distinguish between two scenarios for the proliferation time. 
In the first scenario, the tumor is small, $N\ll N_a^*\ll N_m^*$, and we have at most one lineage that rescues the cancer cell population.
As a result, the recurrence time is given by~\citep{avanzini2019cancer}
\begin{align}\label{meanproliferationtime}
\tau_a^r&\approx\tau_a+\frac{\log \left(p_mN\right)}{\Delta_m}.
\end{align}
The factor $p_m$ in the second term of \cref{meanproliferationtime} occurs because the lineage is conditioned to survive stochastic extinction and the time to reach $N$ is shorter compared to the scenario where the lineage is not conditioned to survive stochastic extinction~\citep{orr2014population,smith1974hitch}. % change 35

In the second scenario, the tumor is large, $N\gg N_m^*$, and the sensitive population produces a large number of rescue lineages in a short period of time.
As a result, the recurrence time is obtained by solving the following system of ordinary differential equations (ODEs),
\begin{equation}\label{detODE}
\begin{aligned}
\frac{ds}{dt}&=\Delta_ss,\\
\frac{da}{dt}&=\Delta_aa+u\lambda_ss,\\
\frac{dm}{dt}&=\Delta_mm+v\lambda_aa+v\lambda_ss.
\end{aligned}
\end{equation}
Solving the system of ODEs for initial condition $\left(s(0), a(0), m(0)\right)=\left(N,0,0\right)$ we obtain:
\begin{equation*} \label{m_t}
m(t) = \frac{N u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m t} - \e^{\Delta_a t}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s} \right] + N v \lambda_s \frac{\e^{\Delta_m t} - \e^{\Delta_s t}}{\Delta_m - \Delta_s}.
\end{equation*}
We obtain $\tau_a^r$ by solving the above equation for time $t=\tau_a^r$ at which the number of mutant cells reaches $N$, that is, we solve $m\left(\tau_a^r\right)=N$,
\begin{equation}\label{eq:t2det2024}
1 = \frac{u v \lambda_a \lambda_s}{\Delta_a - \Delta_s}\left[ \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_a \tau_a^r}}{\Delta_m - \Delta_a} - \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s} \right] + v \lambda_s \frac{\e^{\Delta_m \tau_a^r} - \e^{\Delta_s \tau_a^r}}{\Delta_m - \Delta_s}.
\end{equation}

\Cref{eq:t2det2024} is a transcendental equation which cannot be solved exactly but we can obtain an approximation by noting that the tumor size $N$ is large, and because the sensitive and aneuploid cells have negative growth rates, almost all of the growth needed to rebound to size $N$ will be by mutant cells. 
Mathematically, this corresponds to $\e^{\Delta_m \tau_a^r}$ being much larger than the other exponential terms,  $\e^{\Delta_s \tau_a^r}$ and $\e^{\Delta_a \tau_a^r}$. 
\Cref{eq:t2det2024} can then be rewritten as
\[
	1 \approx \frac{v \lambda_s e^{\Delta_m \tau_a^r}}{\Delta_m - \Delta_s} 
	\left(\frac{u \lambda_a}{\Delta_m - \Delta_a} + 1\right).
\]
Assuming that rate of aneuploidy is not so high that it overwhelms the aneuploids' growth disadvantage relative to the mutants ($u \lambda_a \ll \Delta_m - \Delta_a$), we can neglect the first term in parentheses above, yielding
\begin{equation}\label{eq:t2det}
\tau_a^r\approx\frac{1}{\Delta_m}\log{\left(\frac{\Delta_m-\Delta_s}{v\lambda_s}\right)}.
\end{equation}
We emphasize that in this case the population is large enough to produce mutants without needing to pass through aneuploidy ($N \gg N_m^*$), and so  the terms arising from the evolutionary trajectory \textit{sensitive} $\rightarrow$ \textit{aneuploid} $\rightarrow$ \textit{mutant} do not contribute to the above approximation. %change 9
